Improving Models to Predict Holocellulose and Klason Lignin Contents for Peat Soil Organic Matter with Mid Infrared Spectra

### 2022-03-07 analysis/paper/001-paper-main.Rmd
Update `renv.lock`, update figure format in supporting info for manuscript 1. Update the text. Update the readme.

    To understand global soil organic matter (SOM) chemistry and its dynamics, we need tools to efficiently quantify SOM properties, for example prediction models using mid infrared spectra. The advantages of such models rely on their validity and accuracy. Recently, \cite{Hodgkins.2018} developed models to predict organic matter holocellulose and Klason lignin contents, two indicators of SOM stability and major fractions of organic matter in general. The models can be used to understand large-scale SOM gradients and have been used in various studies.  
    A research gap to fill is that these models have not been validated in detail yet. What are their limitations and how can we improve them? This study provides a validation with the aim to identify concrete steps to improve the models. As a first step, we provide several improvements using the original training data.  
    The major limitation we identified is that the original training data are not representative for a range of peat samples which causes both biased estimates and extrapolation uncertainty under the original models. In addition, the original models can in practice produce unrealistic predictions (negative values or values $>100$ mass-%). Our improved models partly reduce the observed bias, have a better predictive performance for the training data, and avoid unrealistic predictions. Finally, we provide a proof-of-concept that holocellulose contents can also be predicted for mineral-rich samples.  

@@

Mid infrared spectra (MIRS)-based models to predict SOM properties are a promising high-throughput approach which can replace more labor intensive or costly measurements [@ViscarraRossel.2006]. For example, MIRS have been used to predict elemental contents which otherwise are measured using e.g. combustion and gas chromatographic analysis of resulting gases. All the advantages of MIRS-based models rely on the accuracy of the computed models. This makes model validation a crucial step during model development.  

@@

Several models to predict holocellulose and lignin contents in different OM types have been developed (some based on near infrared spectra) [summarized for lignin by @Elle.2019; for holocellulose see e.g. @Peltre.2011; @Sun.2011]. Most of these models consider only wood, material from few species, or specific vegetation organs [@Elle.2019]. Another problem is that most of these models cannot be easily reproduced and used since the raw data has not been published.  
Recently, @Hodgkins.2018 developed models to predict Klason lignin and holocellulose contents from attenuated total reflectance MIRS. The models have several advantages over existing approaches. The training data comprises several different OM types, such as paper products (cardboard, office paper, magazines, newspaper), leaves from diverse species (trees and graminoids), and wood samples [@DelaCruz.2016; @Hodgkins.2018]. Moreover, both raw data and model code are publicly available [@Hodgkins.2018; @Teickner.2020a]. For this reason, the models are particularly suitable for application in other studies [@Teickner.2019; @Moore.2019; @Harris.2020; @Cong.2020; @Cong.2022], most recently in @Verbeke.2022, and for future developments, e.g. by including additional data.   
A problem is that neither @Hodgkins.2018, nor a later study using the models provided a thorough model validation. To compute the models, the authors developed a procedure to extract peaks which are indicative for specific OM fractions, such as holocellulose and Klason lignin (Fig. \@ref(fig:hodgkins-peaks-sample-fig01)), from the MIRS. The peaks are baseline corrected, their maximum height computed and divided by the sum of absorbance values in the spectra. These values are used to calibrate the prediction models. As model validation, only the linearity between the target variables and predictors was assessed (supplementary material to @Hodgkins.2018), but not for example how predicted values match measured values. Moreover, no information has been provided if the training data is representative for SOM, particularly the peat samples analyzed in the study. Similarly, it is unclear if pre-selecting peaks may reduce the predictive accuracy of the models. Many of these concerns have also been raised by an anonymous reviewer of the paper (supplementary information to @Hodgkins.2018). Thus, an important research gap to fill is to validate the models and to provide concrete strategies for further improvements.

@@

Our goals are to identify limitations of the original models and to give concrete recommendations for improvements. Moreover, we use the original data from @Hodgkins.2018 to provide improvements where possible. To this end, we conducted an exploratory analysis using the peat and peat-forming vegetation data provided by @Hodgkins.2018. Our exploratory analysis was guided by the following research questions:

@@

3. _Is the prediction domain of the training data representative for peat samples?_ If a model extrapolates outside the range of predictor variable values in the training data set, it is unclear if the predictions are valid, for instance because it cannot be evaluated if there is bias. An advantage of the independent reference sample set used by @Hodgkins.2018 is that the model is applicable on various OM types and e.g. no problem-specific correlations are used by the models to make predictions [@Hodgkins.2018]. For example, in peat, holocellulose content is controlled by decomposition, and other structures also controlled by decomposition might therefore provide information on holocellulose contents [@Leifeld.2012; @Biester.2014]. But this does not have to be the case for other samples, e.g. peat forming vegetation. An advantage is therefore the generality of the training data set @Hodgkins.2018 used. However, since it does not include peat samples, it is questionable if the prediction domain formed by the training samples covers peat samples and peat forming vegetation in general, and this can result in overlooking prediction failures due to extrapolation. For this reason, we were interested if the training data (its predictions domain, @Wadoux.2021) is representative for peat and peat forming vegetation.

@@

Our aim is to provide a validation of the original models of @Hodgkins.2018 and concrete suggestions for improvements. To this end, we conducted a series of statistical analyzes to query each of our research questions listed in the introduction. During this, we also computed improved models using the same training data that was used to compute the original models. Since we used these improved models to uncover and analyze the limitations of the original models, we first describe steps how we improved the original models and then how we analyzed the limitations.  
We use the data and models collected and developed by @Hodgkins.2018. The training data used to calibrate the original models consist of various organic materials and comprise measured values for holocellulose and Klason lignin [@DelaCruz.2016]. The data are available via @Hodgkins.2018. In addition, we used a data set consisting of MIRS for `r length(unique(d_peat$core_label))` peat cores (`r nrow(d_peat)` samples in total) and `r nrow(d_veg)` vegetation samples collected at the same sites which are also provided by @Hodgkins.2018. We used these peat and vegetation data to evaluate the performance of the original and improved models for peat and peat forming vegetation in general. We assume that the peat samples are representative for a wide range of SOM because they cover a large latitudinal range and a wide range of degrees of decomposition, and were formed by different vegetation [@Hodgkins.2018].  
We computed all models as Bayesian models. This allowed us to consider parameter uncertainty in predictions and to compute models with good predictive performance [@Raftery.2003; @Piironen.2017]. For this reason, we also recomputed the original models as Bayesian models with weakly informative priors to facilitate model comparison using the same metrics and tools. Whenever we refer to "original model", we mean these Bayesian translations, and otherwise refer to "original non-Bayesian models". To show that we did not introduce any of the identified weaknesses using the Bayesian approach, we compared the predictions of the Bayesian models to the original non-Bayesian models (supplementary Fig. 1<!---\@ref(fig:res-p-pred-diff-hol-intensity2)--->).  
All analyzes were performed in R (`r R.Version()$version.string %>% stringr::str_remove("^R ") %>% stringr::str_remove("\\)") %>% stringr::str_replace(" \\(", ", ")`) [@RCoreTeam.2020]. Spectra were preprocessed using ir (`r packageVersion("ir")`) [@Teickner.2020]. The original script used by @Hodgkins.2018 to extract peaks (\url{https://github.com/shodgkins/FTIRbaselines}) as implemented in irpeat (`r packageVersion("irpeat")`) [@Teickner.2020a] was used to recompute the original models and extract peaks from MIRS. Bayesian models were computed using rstanarm (`r packageVersion("rstanarm")`) [@Goodrich.2020] and brms (`r packageVersion("brms")`) [@Burkner.2017; @Burkner.2018] which are interfaces to Stan (`r rstan::stan_version()`) [@StanDevelopmentTeam.2020; @StanDevelopmentTeam.2021]. All predictor variables were $z$-transformed. MCMC sampling was validated in addition using bayesplot (`r packageVersion("bayesplot")`) [@Gabry.2020]. The out-of-sample predictive performance of the models was estimated using the expected log predictive density (ELPD) estimated using Pareto smoothed importance sampling-leave-one-out cross-validation (PSIS-LOO) [@Vehtari.2017] using the loo package (`r packageVersion("loo")`) [@Vehtari.2019]. PSIS-LOO ELPD is an universal measure to compare the predictive performance of models and $-2 \cdot \text{PSIS-LOO ELPD}$ is the deviance [@Vehtari.2017].

@@

Above, we mentioned that assuming a model with normal distribution can result in unrealistic predictions outside the interval [0, 100] mass-%. We were interested if and under which conditions unrealistic predictions can occur. Here, we differentiate two ways in which a prediction can be unrealistic: It can be an unrealistic point estimate (median predictions), and the prediction interval can cover values outside [0, 100] mass-%, even though the median prediction can be in [0, 100] mass-%.  
For our analysis, we first computed a sequence of normalized `carb` and `arom15arom16` peak heights such that the original model predict values completely covering the interval [0, 100] mass-% for holocellulose and Klason lignin, respectively. For these simulated peak heights, we computed median model predictions and 90% prediction intervals using the original models ("original Gaussian models"). Finally, we identified unrealistic predictions in terms of the point estimates and 90% prediction intervals, and the `carb` and `arom15arom16` peak heights under which these occur. We used these values as thresholds to decide if the models make unrealistic predictions in practice. To relate these values to real data, we computed median predictions and 90% prediction intervals also for the peat and vegetation samples and identified unrealistic predictions as for the simulation analysis.  
After identifying conditions under which unrealistic predictions occur, we compared this to the behavior of a model with assumptions in line with the data generating process. Beta regression models assume all values of the dependent variable to be in (0,100) mass-% which is a reasonable assumption given how the actual data is generated. Therefore, we repeated the previous analyses, but now using beta regression models. We used the same model structure and priors for intercept and slopes as the original models ("original beta models").  
To facilitate practical comparison between the modeling approaches, we compared the predictions of all models in the form of depth profiles of the holocellulose and Klason lignin contents. Altogether, this allowed us to identify unrealistic predictions of the original models, and to analyze how the improved models (beta regression models) perform in comparison.

@@

The original model for holocellulose produced unrealistic predictions: It had a negative point prediction for one strongly decomposed sample from a tropical peat core. Moreover, for `r d_peat_pred %>% dplyr::filter(variable_y == "Holocellulose" & model == "Gaussian, Bayesian") %>% dplyr::mutate(negative = lwr < 0) %>% dplyr::summarise(negative = sum(negative)/length(negative)) %>% dplyr::pull(negative) %>% magrittr::multiply_by(100) %>% round(0)`% of the peat samples in the data from @Hodgkins.2018, the lower 90% prediction interval covered negative values (supplementary Fig. 3<!---\@ref(fig:res-p-gaussian-beta-depth-profile)--->). No point prediction or 90% prediction interval was $>100$ mass-%. The original model for Klason lignin did not produce unrealistic predictions for the peat samples (supplementary Fig. 3<!---\@ref(fig:res-p-gaussian-beta-depth-profile)--->). Both models did not produce unrealistic predictions for the vegetation samples.   
Figure \@ref(fig:res-p-gaussian-beta-range-fig02) shows predicted medians and prediction intervals for both models across a potential range of the spectral predictor variables alongside the range covered by the peat and vegetation samples. The model with normal distribution produces negative lower prediction interval limits whenever the predicted median is lower than around `r d_range_pred_limits %>% dplyr::filter(variable_y == "Holocellulose" & variable_limit == "fit_lwr") %>% dplyr::pull(fit) %>% magrittr::multiply_by(100) %>% round(0)` mass-% and upper prediction interval limits $>100$ mass-% whenever the predicted median is larger than around `r d_range_pred_limits %>% dplyr::filter(variable_y == "Holocellulose" & variable_limit == "fit_upr") %>% dplyr::pull(fit) %>% magrittr::multiply_by(100) %>% round(0)` mass-%. For Klason lignin these values are around `r d_range_pred_limits %>% dplyr::filter(variable_y == "Klason lignin" & variable_limit == "fit_lwr") %>% dplyr::pull(fit) %>% magrittr::multiply_by(100) %>% round(0)` and `r d_range_pred_limits %>% dplyr::filter(variable_y == "Klason lignin" & variable_limit == "fit_upr") %>% dplyr::pull(fit) %>% magrittr::multiply_by(100) %>% round(0)` mass-%, respectively.  
The beta regression models produce realistic predictions across the complete range of the spectral predictor variables. In addition, the beta regression model has smaller prediction uncertainties for extreme values (Fig. \@ref(fig:res-p-gaussian-beta-range-fig02)). A consequence of this are larger predicted holocellulose contents with narrower prediction intervals for several (decomposed) peat samples under the beta regression model in comparison to the original normal model (Fig. \@ref(fig:res-p-gaussian-beta-range-fig02) and supplementary Fig. 3<!---\@ref(fig:res-p-gaussian-beta-depth-profile)--->).  
Thus, whereas our results indicate little differences in choosing a model distribution for Klason lignin, for holocellulose contents of peat samples it is crucial to use a beta regression model to avoid unrealistic predictions. Nevertheless, it is generally advisable to use a beta regression model for mass contents. It may not be known in advance how large holocellulose or Klason lignin contents are for a given sample. And even for Klason lignin, contents may be low, e.g. due to high contents of minerals.

@@

Our analysis shows that both strategies to include more variables result in on average more accurate predictions (Table \@ref(tab:res-tab-loo-comparison-res)). This can be concluded based on the average estimated ELPD values (larger values indicate a better estimated predictive accuracy). Consequently, using the data available, we could improve both the model for holocellulose and Klason lignin.  
Is one of the strategies to include more variables advantageous? The models with the largest predictive accuracy are models with small to moderate bin width (Table \@ref(tab:res-tab-loo-comparison-res)). These are better than the respective models using all extracted peaks: in addition to the ELPD, this can be derived from the standard errors of the ELPD differences relative to the best models ($2 \cdot \Delta \text{SE}$ is approximately the 95% confidence interval). In addition, the remaining binned models generally have a better average predictive accuracy than the models using all extracted peaks, except for the model with bin width = 100 cm$^{-1}$ for holocellulose. Summarized, this indicates that --- if a sufficiently high bin resolution (sufficiently small bin width) is chosen --- using binned spectra results in an improved predictive accuracy over the second approach to use all extracted peaks for both holocellulose and Klason lignin.  
To interpret the improved models using binned spectra, we plotted the median coefficients for the best models using binned spectra (Fig. \@ref(fig:res-p-coefficients-fig03)). From this plot, we identified the bins with the largest absolute median coefficients ($\ge0.2$, marked with points and labeled). The reason for the (on average) improved performance is that the improved models can use information contained in the complete spectra which is not contained in the manually selected peaks for the original models:  
Similarly to the original model, a variable near the `carb` peak is important to predict the holocellulose content (bin at ~970 cm$^{-1}$) since it is related to cellulose [@Cocozza.2003; @Stuart.2004; @Artz.2008], but one peak related to aromatic structures which is not extracted with the approach from @Hodgkins.2018 is also important (~1250 cm$^{-1}$: aromatic in plane C-H bending) [@Stuart.2004] (Fig. \@ref(fig:res-p-coefficients-fig03), first panel). This indicates that aromatics provide information on the holocellulose content in the training samples and that extracting only selected peaks may omit useful information. A plausible explanation for the importance of this variable is that holocellulose and Klason lignin together are the major mass fractions in many OM types and both are controlled by the same processes, but often in different directions (decomposition, resource allocation trade-off during plant growth, selective removal during processing) [@Biester.2014; @Chen.2016] (supporting Fig. 4<!---\@ref(fig:res-sup-kl-hol)--->).   
For Klason lignin, the improved model does not use bins near to the `arom15` and `arom16` peaks, in contrast to the original model. Instead bins with large absolute coefficients are located in the fingerprint region and probably related to aromatic in plane C-H bending (~1150 and ~1270 cm$^{-1}$) [@Stuart.2004] (Fig. \@ref(fig:res-p-coefficients-fig03), last panel). We have no explanation why one coefficient is negative and the other positive. In Sect. 3.5<!--"What causes differences in predictions?"-->, we provide a mechanistic interpretation for why the selected variables are important.  
These selected bins are not represented in the extracted peaks. Consequently, models using all extracted peaks cannot use this information and tend to underfit. Similarly, binning with a broad bin width may result in a too coarse spectral resolution, as indicated by the relatively weak predictive accuracy of the model for holocellulose using binned spectra with a bin width of 100 cm$^{-1}$ (Table \@ref(tab:res-tab-loo-comparison-res)).

@@

There were considerable differences in predictions of the original and improved models using binned spectra for both Klason lignin and holocellulose for the peat samples (Fig. \@ref(fig:res-p-pred-diff-all1-fig05) and supplementary Fig. 5 <!---\@ref(fig:res-p-gaussian-beta-depth-profile-best)--->). For holocellulose, the original model tends to predict larger contents for peat, especially for samples with larger holocellulose contents (Fig. \@ref(fig:res-p-pred-diff-all1-fig05)). Similar, albeit less pronounced patterns are visible for the vegetation and training data. For Klason lignin, the improved model predicts up to 25 mass-% larger Klason lignin contents for some samples, but for others up to 10 mass-% smaller contents than the original model. This happens particularly in a region where the original model predicts Klason lignin contents between 25 and 35 mass-% (Fig. \@ref(fig:res-p-pred-diff-all1-fig05)).  
This is surprising for two reasons: First, for the training data, both models make quite similar predictions (Fig. \@ref(fig:res-p-pred-diff-all1-fig05)). The peat samples have `aro15arom16` values similar to the training data (Fig. \@ref(fig:res-p-prediction-domain-fig04)), but nevertheless both models produce contrasting predictions. The same is true for holocellulose for a relatively large fraction of peat samples (Fig. \@ref(fig:res-p-prediction-domain-fig04)). Second, both models have comparatively small bias within the range of the training data relevant to peat samples (supplementary Fig. 6 <!---\@ref(fig:res-p-y-yhat-best)--->). Thus, predictions are different even within the range covered by the training data where the models actually make similar predictions and these differences are not due to misfit of either model to the training data. This indicates that the spectra of the training data are not entirely representative for the peat samples, even if they are within the prediction domain (compare also with the previous section). What causes these differences?

@@

## What can we learn in general from this study?

>>>>

    To understand global soil organic matter (SOM) chemistry and its dynamics, we need tools to efficiently quantify SOM properties, for example prediction models using mid infrared spectra. However, the advantages of such models rely on their validity and accuracy. Recently, \cite{Hodgkins.2018} developed models to quantitatively predict organic matter holocellulose and Klason lignin contents, two indicators of SOM stability and major fractions of organic matter. The models may have the potential to understand large-scale SOM gradients and have been used in various studies.  
    A research gap to fill is that these models have not been validated in detail yet. What are their limitations and how can we improve them? This study provides a validation with the aim to identify concrete steps to improve these models. As a first step, we provide several improvements using the original training data.  
    The major limitation we identified is that the original training data are not representative for a range of diverse peat samples. This causes both biased estimates and extrapolation uncertainty under the original models. In addition, the original models can in practice produce unrealistic predictions (negative values or values $>100$ mass-%). Our improved models partly reduce the observed bias, have a better predictive performance for the training data, and avoid such unrealistic predictions. Finally, we provide a proof-of-concept that holocellulose contents can also be predicted for mineral-rich samples.  

@@

Mid infrared spectra (MIRS)-based models to predict SOM properties are a promising high-throughput approach which can replace more labor intensive or costly measurements [@ViscarraRossel.2006]. For example, MIRS have been used to predict elemental contents which otherwise are measured using e.g. combustion and gas chromatographic analysis of resulting gases. However, all the advantages of MIRS-based models rely on the accuracy of the computed models. This makes model validation a crucial step during model development.  

@@

Several models to predict holocellulose and lignin contents in different OM types have been developed (some based on near infrared spectra) [summarized for lignin by @Elle.2019; for holocellulose see e.g. @Peltre.2011; @Sun.2011]. Most of these models consider only wood, material from few species, or specific vegetation organs [@Elle.2019]. Another problem is that most of these models cannot be easily reproduced and used since the raw data and code have not been published.  
Recently, @Hodgkins.2018 developed models to predict Klason lignin and holocellulose contents from attenuated total reflectance MIRS. The models have several advantages over existing approaches. The training data comprises several different OM types, such as paper products (cardboard, office paper, magazines, newspaper), leaves from diverse species (trees and graminoids), and wood samples [@DelaCruz.2016; @Hodgkins.2018]. Moreover, both raw data and model code are publicly available [@Hodgkins.2018; @Teickner.2020a]. For this reason, the models are particularly suitable for application in other studies [@Teickner.2019; @Moore.2019; @Harris.2020; @Cong.2020; @Cong.2022], most recently in @Verbeke.2022 and @Baysinger.2022, and for future developments, e.g. by including additional data.   
A major problem is that neither @Hodgkins.2018, nor a later study using the models provided a thorough model validation. To compute the models, the authors developed a procedure to extract peaks which are indicative for specific OM fractions, such as holocellulose and Klason lignin (Fig. \@ref(fig:hodgkins-peaks-sample-fig01)), from the MIRS. The peaks are baseline corrected, their maximum height computed and divided by the sum of absorbance values in the spectra. These values are used to calibrate the prediction models. As model validation, only the linearity between the target variables and predictors was assessed (supplementary material to @Hodgkins.2018), but not for example how predicted values match measured values. Moreover, no information has been provided whether the training data is representative for SOM, particularly the peat samples analyzed in the study. Furthermore, it is unclear if pre-selecting peaks may reduce the predictive accuracy of the models. Many of these concerns have also been raised by an anonymous reviewer of the paper (supplementary information to @Hodgkins.2018). Thus, an important research gap to fill is to validate the models and to provide concrete strategies for further improvements.

@@

Our goals are to identify limitations of the original models and to give concrete recommendations for improvements. Moreover, we use the original data from @Hodgkins.2018 to provide improvements on the original models where possible. To this end, we conducted an exploratory analysis using the peat and peat-forming vegetation data provided by @Hodgkins.2018. Our exploratory analysis was guided by the following research questions:

@@

3. _Is the prediction domain of the training data representative for peat samples?_ If a model extrapolates outside the range of predictor variable values in the training data set, it is unclear if the predictions are valid. An advantage of the independent reference sample set used by @Hodgkins.2018 is that the model is applicable on various OM types and e.g. no problem-specific correlations are used by the models to make predictions [@Hodgkins.2018]. For example, in peat, holocellulose content is controlled by decomposition, and other structures also controlled by decomposition might therefore provide information on holocellulose contents [@Leifeld.2012; @Biester.2014]. But this does not have to be the case for other samples, e.g. peat forming vegetation. An advantage is therefore the generality of the training data set @Hodgkins.2018 used. However, since it does not include peat samples, it is questionable if the prediction domain formed by the training samples covers peat samples and peat forming vegetation in general, and this can result in overlooking prediction failures due to extrapolation. For this reason, we were interested if the training data (its predictions domain, @Wadoux.2021) is representative for peat and peat forming vegetation.

@@

We conducted a series of statistical analyzes to query each of our research questions listed in the introduction. During this, we also computed improved models using the same training data that was used to compute the original models. Since we used these improved models to uncover and analyze the limitations of the original models, we first describe steps how we improved the original models and then how we analyzed the limitations.  
We use the data and models collected and developed by @Hodgkins.2018. The training data used to calibrate the original models consist of various organic materials and comprise measured values for holocellulose and Klason lignin [@DelaCruz.2016]. The data are available via @Hodgkins.2018. In addition, we used a data set consisting of MIRS for `r length(unique(d_peat$core_label))` peat cores (`r nrow(d_peat)` samples in total) and `r nrow(d_veg)` vegetation samples collected at the same sites which are also provided by @Hodgkins.2018. We used these peat and vegetation data to evaluate the performance of the original and improved models for peat and peat forming vegetation in general. We assume that the peat samples are representative for a wide range of SOM because they cover a large latitudinal range and a wide range of degrees of decomposition, and were formed by different vegetation [@Hodgkins.2018].  
We computed all models as Bayesian models. This allowed us to consider parameter uncertainty in predictions and to compute models with good predictive performance [@Raftery.2003; @Piironen.2017]. For this reason, we also recomputed the original models as Bayesian models with weakly informative priors to facilitate model comparison using the same metrics and tools. Whenever we refer to "original model", we mean these Bayesian translations, and otherwise refer to "original non-Bayesian models". To check that we did not introduce any of the identified weaknesses using the Bayesian approach, we compared the predictions of the Bayesian models to the original non-Bayesian models (supplementary Fig. 1<!---\@ref(fig:res-p-pred-diff-hol-intensity2)--->).  
All analyzes were performed in R (`r R.Version()$version.string %>% stringr::str_remove("^R ") %>% stringr::str_remove("\\)") %>% stringr::str_replace(" \\(", ", ")`) [@RCoreTeam.2020]. Spectra were preprocessed using ir (`r packageVersion("ir")`) [@Teickner.2020]. The original script used by @Hodgkins.2018 to extract peaks (\url{https://github.com/shodgkins/FTIRbaselines}) as implemented in irpeat (`r packageVersion("irpeat")`) [@Teickner.2020a] was used to recompute the original models and extract peaks from MIRS. Bayesian models were computed using rstanarm (`r packageVersion("rstanarm")`) [@Goodrich.2020] and brms (`r packageVersion("brms")`) [@Burkner.2017; @Burkner.2018] which are interfaces to Stan (`r rstan::stan_version()`) [@StanDevelopmentTeam.2020; @StanDevelopmentTeam.2021]. All predictor variables were $z$-transformed. MCMC sampling was validated in addition using bayesplot (`r packageVersion("bayesplot")`) [@Gabry.2020]. The out-of-sample predictive performance of the models was estimated using the expected log predictive density (ELPD) estimated using Pareto smoothed importance sampling-leave-one-out cross-validation (PSIS-LOO) [@Vehtari.2017] using the loo package (`r packageVersion("loo")`) [@Vehtari.2019]. PSIS-LOO ELPD is an universal measure to compare the predictive performance of models and $-2 \cdot \text{PSIS-LOO ELPD}$ is the deviance [@Vehtari.2017].

@@

Above, we mentioned that assuming a model with normal distribution can result in unrealistic predictions outside the interval [0, 100] mass-%. We were interested if and under which conditions unrealistic predictions can occur. Here, we differentiate two ways in which a prediction can be unrealistic: It can be either an unrealistic point estimate (median predictions), or the prediction interval can cover values outside [0, 100] mass-%, even though the median prediction can be in [0, 100] mass-%.  
For our analysis, we first computed a sequence of normalized `carb` and `arom15arom16` peak heights such that the original model predict values completely covering the interval [0, 100] mass-% for holocellulose and Klason lignin, respectively. For these simulated peak heights, we computed median model predictions and 90% prediction intervals using the original models ("original Gaussian models"). Finally, we identified unrealistic predictions in terms of the point estimates and 90% prediction intervals, and the `carb` and `arom15arom16` peak heights under which these occur. We used these values as thresholds to decide if the models make unrealistic predictions in practice. To relate these values to real data, we computed median predictions and 90% prediction intervals also for the peat and vegetation samples and identified unrealistic predictions as for the simulation analysis.  
After identifying conditions under which unrealistic predictions occur, we compared this to the behavior of a model with assumptions in line with the data generating process. Beta regression models assume all values of the dependent variable to be in (0,100) mass-% which is a reasonable assumption given how the actual data is generated. Therefore, we repeated the previous analyses, but now using beta regression models. We used the same model structure and priors for intercept and slopes as the original models ("original beta models").  
To facilitate practical comparison between the modeling approaches, we compared the predictions of all models in the form of depth profiles of the holocellulose and Klason lignin contents. Altogether, this allowed us to identify unrealistic predictions of the original models, and to analyze how the improved models (beta regression models) perform in comparison.

@@

The original model for holocellulose indeed produced unrealistic predictions: It had a negative point prediction for one strongly decomposed sample from a tropical peat core. Moreover, for `r d_peat_pred %>% dplyr::filter(variable_y == "Holocellulose" & model == "Gaussian, Bayesian") %>% dplyr::mutate(negative = lwr < 0) %>% dplyr::summarise(negative = sum(negative)/length(negative)) %>% dplyr::pull(negative) %>% magrittr::multiply_by(100) %>% round(0)`% of the peat samples in the data from @Hodgkins.2018, the lower 90% prediction interval covered negative values (supplementary Fig. 3<!---\@ref(fig:res-p-gaussian-beta-depth-profile)--->). No point prediction or 90% prediction interval was $>100$ mass-%. The original model for Klason lignin did not produce unrealistic predictions for the peat samples (supplementary Fig. 3<!---\@ref(fig:res-p-gaussian-beta-depth-profile)--->). Both models did not produce unrealistic predictions for the vegetation samples.   
Figure \@ref(fig:res-p-gaussian-beta-range-fig02) shows predicted medians and prediction intervals for both models across a potential range of the spectral predictor variables alongside the range covered by the peat and vegetation samples. The model with normal distribution produces negative lower prediction interval limits whenever the predicted median is lower than around `r d_range_pred_limits %>% dplyr::filter(variable_y == "Holocellulose" & variable_limit == "fit_lwr") %>% dplyr::pull(fit) %>% magrittr::multiply_by(100) %>% round(0)` mass-% and upper prediction interval limits $>100$ mass-% whenever the predicted median is larger than around `r d_range_pred_limits %>% dplyr::filter(variable_y == "Holocellulose" & variable_limit == "fit_upr") %>% dplyr::pull(fit) %>% magrittr::multiply_by(100) %>% round(0)` mass-%. For Klason lignin these values are around `r d_range_pred_limits %>% dplyr::filter(variable_y == "Klason lignin" & variable_limit == "fit_lwr") %>% dplyr::pull(fit) %>% magrittr::multiply_by(100) %>% round(0)` and `r d_range_pred_limits %>% dplyr::filter(variable_y == "Klason lignin" & variable_limit == "fit_upr") %>% dplyr::pull(fit) %>% magrittr::multiply_by(100) %>% round(0)` mass-%, respectively.  
The beta regression models produce realistic predictions across the complete range of the spectral predictor variables. In addition, the beta regression model has smaller prediction uncertainties for extreme values (Fig. \@ref(fig:res-p-gaussian-beta-range-fig02)). A consequence of this are larger predicted holocellulose contents with narrower prediction intervals for several (decomposed) peat samples under the beta regression model in comparison to the original normal model (Fig. \@ref(fig:res-p-gaussian-beta-range-fig02) and supplementary Fig. 3<!---\@ref(fig:res-p-gaussian-beta-depth-profile)--->).  
Thus, whereas our results indicate little differences in choosing a model distribution for Klason lignin, for holocellulose contents of peat samples it is crucial to use a beta regression model to avoid unrealistic predictions. Nevertheless, it is generally advisable to use a beta regression model for mass contents. It may not be known in advance how large holocellulose or Klason lignin contents are for a given sample. And even for Klason lignin, contents may be low, e.g. due to high contents of minerals.

@@

Our analysis shows that both strategies to include more variables result in on average more accurate predictions (Table \@ref(tab:res-tab-loo-comparison-res)). This can be concluded based on the average estimated ELPD values (larger values indicate a better estimated predictive accuracy). Consequently, using the data available, we could improve both the model for holocellulose and the model for Klason lignin.  
Is one of the strategies to include more variables advantageous? The models with the largest predictive accuracy are models with small to moderate bin width (Table \@ref(tab:res-tab-loo-comparison-res)). These are better than the respective models using all extracted peaks: in addition to the ELPD, this can be derived from the standard errors of the ELPD differences relative to the best models ($2 \cdot \Delta \text{SE}$ is approximately the 95% confidence interval). In addition, the remaining binned models generally have a better average predictive accuracy than the models using all extracted peaks, except for the model with bin width = 100 cm$^{-1}$ for holocellulose. Summarized, this indicates that --- if a sufficiently high bin resolution (sufficiently small bin width) is chosen --- using binned spectra results in an improved predictive accuracy over the second approach to use all extracted peaks for both holocellulose and Klason lignin.  
To interpret the improved models using binned spectra, we plotted the median coefficients for the best models using binned spectra (Fig. \@ref(fig:res-p-coefficients-fig03)). From this plot, we identified the bins with the largest absolute median coefficients ($\ge0.2$, marked with points and labeled). The reason for the (on average) improved performance is that the improved models can use information contained in the complete spectra which is not contained in the manually selected peaks for the original models:  
Similarly to the original model, a variable near the `carb` peak is important to predict the holocellulose content (bin at ~970 cm$^{-1}$) since it is related to cellulose [@Cocozza.2003; @Stuart.2004; @Artz.2008], but one peak related to aromatic structures which is not extracted with the approach from @Hodgkins.2018 is also important (~1250 cm$^{-1}$: aromatic in plane C-H bending) [@Stuart.2004] (Fig. \@ref(fig:res-p-coefficients-fig03), first panel). This indicates that aromatics provide information on the holocellulose content in the training samples and that extracting only selected peaks may omit useful information. A plausible explanation for the importance of this variable is that holocellulose and Klason lignin together are the major mass fractions in many OM types and both are controlled by the same processes, but often in different directions (decomposition, resource allocation trade-off during plant growth, selective removal during processing) [@Biester.2014; @Chen.2016] (supporting Fig. 4<!---\@ref(fig:res-sup-kl-hol)--->).   
For Klason lignin, the improved model does not use bins near to the `arom15` and `arom16` peaks, in contrast to the original model. Instead bins with large absolute coefficients are located in the fingerprint region and probably related to aromatic in plane C-H bending (~1150 and ~1270 cm$^{-1}$) [@Stuart.2004] (Fig. \@ref(fig:res-p-coefficients-fig03), last panel). We have no explanation for why one coefficient is negative and the other positive. In Sect. 3.5<!--"What causes differences in predictions?"-->, we provide a mechanistic interpretation for why the selected variables are important.  
The bins with especially large absolute coefficients in the models using binned spectra are not represented in the extracted peaks because these cover different wavenumber ranges (compare Fig. \@ref(fig:res-p-coefficients-fig03) to Fig. \@ref(fig:hodgkins-peaks-sample-fig01)). Consequently, models using all extracted peaks cannot use this information and tend to underfit. Similarly, binning with a broad bin width may result in a too coarse spectral resolution, as indicated by the relatively weak predictive accuracy of the model for holocellulose using binned spectra with a bin width of 100 cm$^{-1}$ (Table \@ref(tab:res-tab-loo-comparison-res)).

@@

There were considerable differences in predictions of the original and improved models using binned spectra for both Klason lignin and holocellulose for the peat samples (Fig. \@ref(fig:res-p-pred-diff-all1-fig05) and supplementary Fig. 5 <!---\@ref(fig:res-p-gaussian-beta-depth-profile-best)--->). For holocellulose, the original model tends to predict larger contents for peat, especially for samples with larger holocellulose contents (Fig. \@ref(fig:res-p-pred-diff-all1-fig05)). Similar, albeit less pronounced patterns are visible for the vegetation and training data. For Klason lignin, the improved model predicts up to 25 mass-% larger Klason lignin contents for some samples, but for others up to 10 mass-% smaller contents than the original model. This happens particularly in a region where the original model predicts Klason lignin contents between 25 and 35 mass-% (Fig. \@ref(fig:res-p-pred-diff-all1-fig05)).  
This is surprising for two reasons: First, for the training data, both models make quite similar predictions (Fig. \@ref(fig:res-p-pred-diff-all1-fig05)). The peat samples have `aro15arom16` values similar to the training data (Fig. \@ref(fig:res-p-prediction-domain-fig04)), but nevertheless both models produce contrasting predictions. The same is true for holocellulose for a relatively large fraction of peat samples (Fig. \@ref(fig:res-p-prediction-domain-fig04)). Second, both models have comparatively small bias within the range of the training data relevant to peat samples (supplementary Fig. 6 <!---\@ref(fig:res-p-y-yhat-best)--->). Thus, predictions are different for the peat samples even within the holocellulose and Klason lignin range covered by the training data, where the models actually make similar predictions for the training data. And these differences are not due to misfit of either model to the training data. This indicates that the spectra of the training data are not entirely representative for the peat samples, even if they are within the prediction domain (compare also with the previous section). What causes these differences?

@@

## General implications of our results

Our validation analysis provides general lessons for validating models using spectral data. What can we learn from the model validation? First, even though @Hodgkins.2018 had a sensible rationale in including only selected peaks into their model based upon causal knowledge, this strategy has critical weaknesses. If predictive accuracy is the main goal, pre-selecting variables triggers underfitting and bias and therefore is ineffective. Instead, it is more effective to include a larger set of variables in a model and to use regularization to avoid overfitting and find relations between predictor variables and the target variable [@Piironen.2017].  
